### 2009-05-27 response-to-reviews2.txt

We do hope that the sophistication of the constraint schemes used with
these "tree transducer" models will increase over time. Eventually we
hope to develop alignment constraints, simultaneous alignment-and-fold
constraints, and other constraints (and/or sampling schemes) to enable
efficient analysis of multiple phylogenetically-related sequences. The
present work represents a first attempt at using these sorts of models
for more than two sequences, but we hope it will not be the last
attempt, and we acknowledge room for improvement.

> It is still not clear eenough what the authors' model is, though it is 
> supposed to be a main result. Page 7 states "a prescriptive formalism 
> for transducers is given in Text S1". Relegating the deliverables of a 
> manuscript to supplementary material is not an appropriate use of 
> supplementary material; a reader should not have to refer to a second 
> manuscript to understand the main points. The authors need to find a 
> way to explain their main results in the actual manuscript. There is 
> still much vagueness and redundancy in this draft, which could be 
> removed and replaced with more relevant explanation. For example, in 
> this same section where results are relegated to supplementary 
> material, there's a whole redundant paragraph (p.8) on how conditional 
> models represent directed edges; but this concept has already been 
> explained sufficiently. This is followed (bottom p. 8, "Consider a 
> phylogenetic tree...") by yet another statement of the same concept of 
> using conditional models, still in
> terms too vague and conceptual to enable reproduction of the work.

Thank you for pointing this out.  We have removed the reference to the
supplementary material in this section and have attempted to make the
main text a self-contained description of the composition algorithm,
including its key corollaries (elimination of null states, and DP inference).
We have attempted to remove redundant text, including the two
paragraphs which the reviewer notes, and other sentences.

@@

> It is still not clear enough what the authors mean by a "composition 
> algorithm". The description of the generative model on pp 9-10 seems 
> redundant with the "composition algorithm" on pp 11-12; pp 11-12 don't 
> seem to add anything that I didn't already understand from pp 9-10, 
> and is still not precise and concrete enough for me to really 
> understand the actual model in play here. My understanding is that a 
> "composition algorithm" is just how one chains conditional models 
> together to calculate the desired joint probability of the observed 
> data. Conceptually, this chaining seems obvious enough to me when a 
> phylogenetic tree is given, and the use of so much space to describe 
> the concept is puzzling; but perhaps I'm missing something. In any 
> case, what the reader really wants is a precise explanation of the 
> exact model in use here, not a lot of text about the well-understood 
> concept of conditional likelihoods on branches of a tree.

We have reorganized the presentation of the "composition algorithm" to
attempt to clarify both what it is and why it is important.  The
reviewer is completely correct that it is, indeed, a prescription
for chaining together conditional models.  However, while the concept
of so chaining together models is intuitive, in practice it can be a
daunting task for SCFGs thanks to the presence of bifurcations.  For
example, while both the model generating a prior distribution over ancestral 
sequences (the "singlet transducer," with 7 states) and the model describing
evolution along a single branch (the "branch transducer," with 21 states)
are simple and compact, the three-taxon ensemble model has 230 states
and 1,789 transitions.

In summary, while we completely agree that the model construction procedure
is no more than a formalization of an intuitive concept, we believe
that it is nonetheless an essential reference point, due to the sheer complexity
of the ensemble models so created.  The "TKFST" model which we use
here is relatively simple; more realistic models will give rise to even
more complex ensemble models.

@@

We agree that these reconstructions were not very informative.
Unfortunately these were the only sequences that we could find which
were short enough to be tractable on the hardware we had available. We
now report simulation results instead, as suggested.

We thank the reviewer for pointing out errors in the original
figures. These were due to mistakes made during manual conversion of
indiegram's output parse trees into Stockholm-format alignments and a
bug in the code used for maximum-likelihood reconstruction of
sequences (the xrate program). We have now automated this process and
fixed the bug, verifying that this has eliminated the errors
(in fact this verification is incorporated into our new benchmark).

@@

> p.6 bottom: pair HMMs are not really "standard"; they are not widely 
> used or known. (There is still too much of a general tendency in the 
> manuscript to use technically narrow jargon in place of actual 
> explanation, and assume the reader will be able to fill in the gaps. 
> Sometimes, though, the revision gets it right: for example, other than 
> this "standard pair HMM" reference, the authors do an excellent job on 
> pp6-7 introducing the concept and motivation of conditional 
> probability models first, and *then* connecting to the linguistics 
> literature and the jargon term "parse-tree transducer".)

Thank you for pointing out this oversight.  We briefly describe what a
Pair HMM is in the revised manuscript.

> p.7 why is the model that generates P(X) at the root a "singlet 
> transducer"? It's not transducing anything (it's not conditionally 
> dependent on an input parse tree).

By "singlet transducer" we mean the same thing as is often referred to
as a "single-sequence SCFG".  Instead of "a model that generates a
parse tree X", we use the contrivance of saying "a model that inputs a
(dummy) empty parse tree, and outputs a parse tree X".  It's annoying
but it keeps the notation straight: it allows us to be consistent
throughout the manuscript. Rather than always having to specify
whether a model is generative (SCFG) or transductive (input-output),
we can treat all models as transducers, if we just think of generative
models as transducers with a (dummy) input that is always zero/empty.

We apologize if it is poor terminology. We just think it's as clear as
we can get. We can change it if the reviewer feels that doing so would
help clarify the methods.

### 2009-05-27 paper.tex
cut swathes out of intro to elimination/DP supersection

In this section, we describe dynamic programming (DP) algorithms for
inferring the alignment, structure and evolutionary history
of multiple related RNAs, using the multiple-sequence SCFG we have derived.
The model construction algorithm described above creates an evolutionary model
which generates a joint distribution over both ancestral and extant
sequences.  If we regard the ancestral sequences as hidden variables
to be marginalized over, then this model corresponds to a
jointly-normalized multiple-sequence SCFG (modeling the extant sequences).
We can then use the Cocke-Younger-Kasami (CYK) algorithm to infer the
maximum-likelihood alignment, secondary structure and common ancestor
of the extant sequences or the Inside algorithm to sample from
the joint posterior distribution of these unknowns.

The possibility of fragments of ancestral sequence that are unobserved
in any of the descendant sequences, which can be unbounded in length,
leads to complications when performing inference on the multiple-sequence
SCFG. The SCFG can contain null cycles, corresponding to generation of
these unobserved fragments, which complicates inference.

Due to this possibility of unobserved ancestral sequence,
the standard CYK and Inside recursions fail on the ensemble model:
that is, they contain cyclic self-references which cannot be resolved.
The intuitive reason for this is that a (potentially-unbounded) number of residues in the ancestor can be
absent in all extant sequences.
This is manifested as cycles of $\Snull$ states within the multiple-sequence SCFG,
states which emit to the ancestral sequence but not to any extant sequence.
The presence of bifurcations with one or more possibly-empty child branches,
corresponding to entire substructures in the ancestral sequence that
are absent in any extant sequence, further complicates inference.
While there exists a straightforward fix for the CYK algorithm, the Inside algorithm is more complicated.

We consider the CYK algorithm first.
In order to correctly perform maximum-likelihood inference, we only
require that the maximum-likelihood parse tree corresponds to the parse
returned by the CYK algorithm; the entire
probability distribution over parses need not be preserved.
Because a state path which includes a cycle of $\Snull$ states
necessarily has lower probability than an equivalent one without, we
can simply ignore such $\Snull$ cycles during CYK.
Bifurcations with empty child branches cannot be similarly ignored,
but this problem can be addressed by creating equivalent effective
direct transitions between the parents of these bifurcation states and
their non-empty children.  Once these transformations have been performed,
maximum-likelihood inference can be performed using a standard
three-sequence CYK algorithm on the ensemble model.
(Similarly, the null cycle elimination discussed in computational
linguistics requires only that the CYK parse be preserved \cite{GecsegSteinby97}.)

The transformations just described to make CYK work properly are insufficient for the
Inside and Outside algorithms.
For CYK, parse trees containing null cycles can be ignored, since the maximum likelihood parse tree should never contain null cycles.
For Inside and Outside, such null-cycle parses cannot be ignored, since they will contribute to the likelihood sum over parse trees.

There are two possible solutions to this difficulty: we can either
create new versions of the standard inference algorithms for SCFGs
which can cope with SCFGs with null cycles or we can modify the
structure of the SCFGs generated by our model-construction algorithm
such that they are free of null cycles.  The first approach of
creating new algorithms is simple in principle---we can simply
iteratively re-fill entries of the DP matrix, corresponding to summing
the probabilities for unobserved null cycles, until the approximation
is sufficiently good---but it is difficult to implement efficiently
and necessarily has worse time complexity than the standard inference
algorithms due to the need to reiterate over (large) parts of the DP
matrix.  The second approach, of modifying the structure of the SCFGs
themselves, is well-understood for the case of HMMs, but to our
knowledge, still an open problem for SCFGs
(complicated by the presence of null cycles involving bifurcations).

We therefore developed a method to eliminate null cycles from
the ensemble model, in order to use the standard inference algorithms
for SCFGs without incurring the further computational complexity
required by modifying the DP algorithms themselves.
We will describe this elimination procedure before proceeding to the descriptions of the DP algorithms.

@@

We refer to empty paths and empty bifurcations collectively as {\bf null cycles}.
Our goal is an algorithmic procedure to account for the likelihood of such cycles by exact marginalization.
For simpler models, solutions to this problem are published.
Eliminating missing (empty) columns from an independent-sites model of sequence involves a straightforward application of a correction factor
to account for the proportion of columns that are unobserved \cite{RivasEddy2008}.
The slightly more complicated situation of missing emissions in a HMM can be dealt with by
summing over all empty paths, yielding a geometric series that is solvable by matrix inversion
\cite{Holmes2003,Holmes2007,Lunter2007}.

@@

Once we have performed the transformations described above to remove
null cycles from the multiple-sequence SCFGs generated by our
model-construction algorithm, we can use the standard CYK, Inside and
Outside algorithms for multiple-sequence SCFGs given by Sankoff \cite{Sankoff85}.

The asymptotic time and memory complexities of our inference algorithms
are essentially the same as for the Sankoff algorithm \cite{Sankoff85}:
the DP algorithms take memory $O(A \cdot L^{2N})$ and time $O(B \cdot L^{3N})$
for $N$ sequences of length $L$, where $A$ is the number of
(accessible) states in the multiple-SCFG and $B$ is the number of
bifurcations.
Note that $A$ and $B$ are also dependent on $N$ (see ``The TKFST model
on a three-taxon phylogeny'').

Exact inference on a star phylogeny with $N$ extant sequences therefore
has complexities $O(A \cdot L^{2N})$ and $O(B \cdot L^{3N})$ in space and
time for a multiple-SCFG with $A$ states and $B$ bifurcations. As
described earlier, in practice we frequently have expert knowledge
about the structures or evolutionary histories of the sequences of
interest, such as a curated multiple alignment.  We can use these
constraints to reduce the accessible volume of the DP matrix \cite{Holmes2005}.
Algorithms \ref{alg:inside}--\ref{alg:cyktraceback}, described below, give the Inside, CYK,
Outside and CYK traceback algorithms for a three-taxon star phylogeny,
constrained using the ``fold envelope'' concept.

>>>>

In this section, we describe dynamic programming (DP) algorithms for
inferring the alignment, structure and evolutionary history
of multiple related RNAs, using the multiple-sequence SCFG we have derived.

The model construction algorithm described above creates a phylogenetic SCFG
which generates a joint distribution over both ancestral and extant
sequences. A parse tree for this SCFG represents a structural and evolutionary
explanation of the extant sequences, including a complete ancestral reconstruction.
Therefore, given a set of extant sequences, many of the questions of interest to us
can be reduced to searches over, or summarizations of, the set of possible parse trees.

Well-known algorithms exist for searching or summing over SCFG parse trees.
The Cocke-Younger-Kasami (CYK) algorithm performs maximum-likelihood (ML) inference;
the Inside algorithm can be used to sum over parse trees or sample them {\em a posteriori};
and the Inside-Outside algorithm yields posterior probabilities for individual parse tree nodes.

All of these algorithms are, however, complicated for phylogenetic SCFGs by
the possibility of fragments of ancestral sequence that were universally deleted and therefore
are unobserved in any of the extant sequences.
These unobserved fragments can be unbounded in length.
In the parse tree, these fragments correspond to cyclic detours through $\Snull$ states,
along with bifurcations where one or both sides terminate without observable emissions
(e.g. $B \to X\ Y$ followed by $X \to \epsilon$).
In the CYK, Inside and Outside recursions, this causes cyclic dependencies which cannot be resolved.

Therefore, before presenting equivalents of the CYK, Inside and Outside algorithms,
we first give a method to eliminate ``null cycles'' from
the ensemble model, by transforming any SCFG to an equivalent acyclic SCFG.
Some such step is required, in order to deal with cyclic dependencies, but alternative approaches to ours are clearly possible.
A CYK-only approach appears in the computational linguistics literature \cite{GecsegSteinby97}.
We have also developed an independent heuristic that iteratively re-fills entries of the DP matrix
until the approximation is sufficiently good.

@@

Empty paths and empty bifurcations can be combined to give finite-probability sequences of rules
that transform a nonterminal back into itself, with no observable emissions.
We refer to such sequences of rules as {\bf null cycles}.
Our goal is an algorithmic procedure to account for the likelihood of such cycles by exact marginalization.
For simpler models, solutions to this problem are published.
Eliminating missing (empty) columns from an independent-sites model of sequence involves a straightforward application of a correction factor
to account for the proportion of columns that are unobserved \cite{RivasEddy2008}.
The slightly more complicated situation of missing emissions in a HMM can be dealt with by
summing over all empty paths, yielding a geometric series that is solvable by matrix inversion
\cite{Holmes2003,Holmes2007,Lunter2007}.

@@

Once we have performed the transformations described above to remove
null cycles from the multiple-sequence SCFGs generated by our
model-construction algorithm, we can compute likelihoods and sample parse trees
using the ``standard'' CYK, Inside and Outside algorithms for multiple-sequence SCFGs \cite{Sankoff85,Durbin98}.

The asymptotic time and memory complexities of our inference algorithms
are essentially the same as for Sankoff's algorithm \cite{Sankoff85}:
the DP algorithms take memory $O(A \cdot L^{2N})$ and time $O(B \cdot L^{3N})$
for $N$ sequences of length $L$, where $A$ is the number of
(accessible) states in the multiple-SCFG and $B$ is the number of
bifurcations.
Note that $A$ and $B$ are also dependent on $N$ (see ``The TKFST model
on a three-taxon phylogeny'').

Exact inference on a star phylogeny with $N$ extant sequences therefore
has complexities $O(A \cdot L^{2N})$ and $O(B \cdot L^{3N})$ in space and
time for a multiple-SCFG with $A$ states and $B$ bifurcations. As
described earlier, in practice we frequently have expert knowledge
(such as a curated multiple alignment)
about the structures and/or evolutionary histories of the sequences of
interest.  We can use this knowledge as a constraint to reduce the accessible volume of the DP matrix \cite{Holmes2005}.
Algorithms \ref{alg:inside}--\ref{alg:cyktraceback}, described below, give the Inside, CYK,
Outside and CYK traceback algorithms for a three-taxon star phylogeny,
constrained using the ``fold envelope'' concept.
